### 2015-05-25 submission20150525/plos_template.tex
organized directory

Our methods are implemented in an R package 
pmsignature (\url{https://github.com/friend1ws/pmsignature})
and a web application available at \url{https://friend1ws.shinyapps.io/pmsignature_shiny/}.

@@

Software implementing the proposed methods is available in an R package 
{\bf pmsignature} ({\bf p}robabilistic {\bf m}utation signature),
at \url{https://github.com/friend1ws/pmsignature}).
The core part of the estimation process is implemented in C++ by way of the Rcpp package \cite{eddelbuettel2011rcpp},
which enables handling millions of somatic mutations from thousands of cancer genomes using a standard desktop computer.
In addition, a web-based application of our method is available at 
\url{https://friend1ws.shinyapps.io/pmsignature_shiny/}

% Results and Discussion can be combined.
\section*{Methods Overview}

@@

\caption{{\bf Examples of visualizations and parameter values for the mutation signatures of the unconstrained (full) model and our independent model, 
where substitution patterns, two 5' and 3' bases and transcription strand direction are considered as mutation features.}
(A) The barplots are divided by 6 substitution patterns and transcription strand direction.
In each division, 256 bars show joint probabilities of up to two base 5' and 3' bases with substitution patterns and transcription strands
(ApApNpApA, ApApNpApC, ApApNpApG, ApApNpApT, $\cdots$, TpTpNpTpT).
(B, C) An example mutation signature representation and parameter values from our independent model, where mutation features (substitution patterns, two 5' and 3' bases and strand direction) 
are assumed to be independent ($L=6$, $\bm{M} = (6, 4, 4, 4, 4, 2)$).
In the bottom five rectangles, the width of each box represents the frequencies of bases (A, C, G and T) at the substitution and flanking site. 
To highlight the most informative flanking sites, the heights of flanking site boxes are scaled by ($1 + 0.5 \times \log \sum_{n = A,C,G,T} f_n^2$, where $f_n$ is the parameter for each base), which can be interpreted as ($1 - 0.5 \times \text{R\'enyi entropy}$) \cite{rrnyi1961measures}. This is analogous to the information content scaling used in sequencing logos.
In the top rectangle, the height of each box represents the conditional frequencies of mutated bases for each original base (C and T).
In the upper right, the height of the $+$ box represents the frequencies of mutations in the coding strand (or the plus strand, the sense strand and the untranscribed strand)
whose nucleotide sequences directly corresponds to mRNA, 
whereas the height of $-$ box represents those in the template strand (or the minus strand, the antisense strand, the transcribed strand and the noncoding strand)
whose sequences are copied during the synthesis of mRNA.
}
\label{mutSig_example}

@@

\subsection*{Mathematical specification, and connection with mixed-membership models}
% yshira20150421 I think the title of the section should change (because this is not "detail").

@@

This generative model is summarized in Fig. \ref{methodOverview}.
The key parameters in this model
are the membership proportions for each sample, $\bm{q}_i$, and the mutation signature parameters, $\bm{F}_k$. 
We estimate these parameters by maximizing likelihood using an EM algorithm. 
A simulation study demonstrates that the estimation method can reproduce the mutation signature very accurately 
provided enough mutations and samples are available (see S1 Text). 
See Methods for more details, how to select $K$, etc.

\begin{figure}[h]
% \includegraphics[width=450pt]{methodOverview.pdf}
\caption{{\bf An overview of the generative model of somatic mutations proposed in this paper.} 
Suppose there are three types of mutation sources (mutation signatures) such as ultraviolet, tobacco smoking chemicals and transcription coupled repairs.
Each cancer genome has ratios showing which types of mutation sources are contributing to its mutations (membership parameters).
The generative model of the pattern of each mutation is:
first, one of the mutation signatures is chosen according to the membership parameter.
Second, according to the selected mutation signature, mutation features such as substitution pattens and flanking bases are generated 
by the corresponding multinomial distributions for that mutation signature.
}
\label{methodOverview}
\end{figure}

@@

This model is essentially a ``mixed-membership model", also known as an ``admixture model" \cite{pmid10835412} or ``Latent Dirichlet Allocation" \cite{Blei:2003}. For example,
the membership proportions for each sample are analogous to admixture proportions in an admixture model; the mutation signatures are analogous to 
populations, and the mutation signature-specific parameters are analogous to population-specific allele frequencies
(see Relationship with mixed-membership models in Methods for further discussion).

% Results and Discussion can be combined.
\section*{Results}

@@

Here we compare our new ``independent model'' for 
mutation signatures, which assumes independence
among mutation features,
with the ``full model'', which corresponds to existing approaches. We compare mutation signatures obtained by the two approaches and investigate the robustness of each approach by downsampling experiments.

@@

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and results in the next subsection ((Figs. 3A and 3B)).
In contrast, for the full model, this tendency is rather mild (Figs. 3C, 3D and S1).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

\begin{figure}[h]
\caption{{\bf The mutation signatures for the UCUT data, and the results of downsampling experiments.}
3072 elements in the full model mutation signatures were shown divided by 6 substitution patterns and strand directions.
(A) APOBEC signature for the independent model.
(B) AA signature for the independent model.
(C) APOBEC signature for the full model.
(D) AA signature for the full model.
(E) APOBEC signature stability.
(F) AA signature stability.}
\label{UCUT}
\end{figure}

To investigate this further we performed downsampling experiments.
Using the mutation signatures obtained using all 14717 substitutions as a gold standard, 
we assessed performance of the proposed method on down-sampled data consisting of $r\%$ of the original data, where  $r=(1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%)$.
To measure robustness we used the cosine similarity on the 
full dimensional vector space, which allows comparison between the full model and the independent model.
We repeated each down-sampling experiment 100 times for each model.

@@

\begin{figure}[h]
\centering
% \includegraphics[width=16cm,height=17.5cm]{AlexandrovEtAl_mergedSignature.eps}
\caption{{\bf The summary of mutation signatures obtained in a reanalysis of the data of the previous study \cite{pmid23945592} using the proposed method.}
Here, the substitution patterns and two 5' and 3' bases from the mutated sites are taken into account as mutation features.
First, mutation signatures were estimated separately in each cancer type, and then similar signatures were merged (see text).
}
\label{nature2013_sig_summary}
\end{figure}

@@

\begin{figure}[h]
% \subfigure[APOBEC signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=14.5cm,height=6cm]{APOBEC_two5prime.eps}
%    \label{APOBEC_two5prime}}
% \subfigure[POLE1 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE1_two5prime.eps}
%   \label{POLE1_two5prime}}
% \quad
% \subfigure[POLE2 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE2_two5prime.eps}
%   \label{POLE2_two5prime}}
% \subfigure[UV signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{UV_two5prime.eps}
%   \label{UV_two5prime}}
% \quad
% \subfigure[The signature 11 intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{LMST_two5prime.eps}
%   \label{LMST_two5prime}}
\caption{{\bf The estimated frequencies of bases at two 5' to the mutated site for each cancer type.}
The bar heights show the estimated frequency for bases A, C, G and T at two 5' to the mutated site.
The error bars show bootstrapped standard errors.
(A) The signature 13 (APOBEC signature)  intensities at two 5' to the mutated site.
(B) The signature 1 (the first POL $\epsilon$ signature) intensities at two 5' to the mutated site.
(C) The signature 8 (the second POL $\epsilon$ signature) intensities at two 5' to the mutated site.
(D) The signature 10 (ultraviolet signature) intensities at two 5' to the mutated site.
(E) The signature 11 intensities at two 5' to the mutated site.
}
\label{two5prime}
\end{figure}

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'' \cite{pmid23318258}). Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S3A and S3B), and signature 2 disappeared when we removed this sample (see Fig. S4).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

@@

We have detected a number of mutation signatures having transcription strand biases, 
which are naturally considered to be associated with transcription activities.
Therefore, to further understand the effect of transcription activities on mutational mechanisms, 
our framework will be helpful: by allowing for including gene expression or RNA polymerase II occupancies 
to mutation features, the relationships of strand biases and transcription activities will be clarified.

@@

We use the R package SQUAREM \cite{varadhan2008simple} to accelerate
convergence of this EM algorithm (SQUAREM implements a general approach to accelerate the convergence of any fixed-point iterative scheme such as an EM algorithm).
To address potential problems with convergence to local minima,
we apply the EM algorithm several times (10 times in this paper) using different initial points,
and use the estimate with the largest log-likelihood.
See S3 for the derivation of the above updating procedures.

@@

\subsection*{S1 Table}
\label{mfexample}
{\bf Example of representation for mutation patterns (substitution patterns and one 5' and 3' bases)}
In the independent representation, the elements of vector show substitution patterns, 5' adjacent bases and 3' adjacent bases, respectively.
For substitution pattens, 1 to 6 values are assigned to C$>$A, C$>$G, C$>$T, T$>$A, T$>$C and T$>$G in this order.
For 5' and 3' adjacent bases, 1 to 4 values are assigned to A, C, G and T.
Note that the original base is fixed to C or T to remove the redundancy of complement sequences.

\subsection*{S2 Table}
\label{nature2013_sigNum}
{\bf The number of mutation signatures selected for each cancer type.}

\subsection*{S1 Figure}
\label{UCUT_APOBEC_twoFivePrime}
{\bf The frequencies of bases at two 5' to the mutated site for the APOBEC mutation signatures obtained in UTUC data using the independent and full models.}

\subsection*{S2 Figure}
\label{nature2013_sig_list}
{\bf The list of several signatures extracted in each cancer type in the Alexandrov et al. (2013) data.}
(A) APOBEC signatures obtained in each cancer type.
(B) Smoking signature in each cancer type.
(C) The first Pol $\epsilon$ signature (signature 1)  in each cancer type.
(D) The second Pol $\epsilon$ signature (signature 8) in each cancer type.
(E) The ultraviolet signature in each cancer type.
(F) Unknown signature obtained in lung small cell carcinomas and stomach cancers.

@@

\subsection*{S4 Figure}
\label{LGG_signatures}
{\bf The signatures obtained for the original data and for the data without the hyper-mutated case.}
(A) The result for the original data ($K = 3$). The first (from the left) signature seems to be deamination of 5-methyl-cytosine. 
  The second signature is the low grade glioma signature.
(B, C, D, E) The result for the data without the hyper-mutated sample for $K = 2, 3, 4$ and $5$, respectively.
Although the signature related to deamination of 5-methyl-cytosine remained, low grade glioma specific signature could not be observed.

@@

We thank to Dr. Daichi Mochihashi for helpful discussion and comments on an earlier version of the method;
Dr. Hiromichi Suzuki for helpful feedback on the R packages;
Dr. Ziyue Gao for helpful comments on the manuscript.

>>>>

Our methods are implemented in an R package 
{\bf pmsignature} (\url{https://github.com/friend1ws/pmsignature})
and a web application available at \url{https://friend1ws.shinyapps.io/pmsignature_shiny/}.

@@

Software implementing the proposed methods is available in an R package 
{\bf pmsignature} ({\bf p}robabilistic {\bf m}utation signature),
at \url{https://github.com/friend1ws/pmsignature}.
The core part of the estimation process is implemented in C++ by way of the Rcpp package \cite{eddelbuettel2011rcpp},
which enables handling millions of somatic mutations from thousands of cancer genomes using a standard desktop computer.
In addition, a web-based application of our method is available at 
\url{https://friend1ws.shinyapps.io/pmsignature_shiny/}.

% Results and Discussion can be combined.
% \section*{Methods Overview}
\section{Result}

@@

\caption{{\bf Examples of visualizations and parameter values for the mutation signatures of the unconstrained (full) model and our independent model, 
where substitution patterns, two 5' and 3' bases and transcription strand direction are considered as mutation features.}
(A) The barplots are divided by 6 substitution patterns and transcription strand direction.
In each division, 256 bars show joint probabilities of up to two base 5' and 3' bases
(ApApNpApA, ApApNpApC, ApApNpApG, ApApNpApT, $\cdots$, TpTpNpTpT).
(B, C) An example mutation signature representation and parameter values from our independent model, where mutation features (substitution patterns, two 5' and 3' bases and strand direction) 
are assumed to be independent ($L=6$, $\bm{M} = (6, 4, 4, 4, 4, 2)$).
In the bottom five rectangles, the width of each box represents the frequencies of bases (A, C, G and T) at the substitution and flanking site. 
To highlight the most informative flanking sites, the heights of flanking site boxes are scaled by ($1 + 0.5 \times \log \sum_{n = A,C,G,T} f_n^2$, where $f_n$ is the parameter for each base), which can be interpreted as ($1 - 0.5 \times \text{R\'enyi entropy}$) \cite{rrnyi1961measures}. This is analogous to the information content scaling used in sequencing logos.
In the top rectangle, the height of each box represents the conditional frequencies of mutated bases for each original base (C and T).
In the upper right, the height of the $+$ box represents the frequencies of mutations in the coding strand (or the plus strand, the sense strand and the untranscribed strand)
whose nucleotide sequences directly corresponds to mRNA, 
whereas the height of $-$ box represents those in the template strand (or the minus strand, the antisense strand, the transcribed strand and the noncoding strand)
whose sequences are copied during the synthesis of mRNA.
}
\label{mutSig_example}

@@

% \subsection*{Mathematical specification, and connection with mixed-membership models}
\subsection*{An overview of mathematical specification of mutation signatures and the generative model of somatic mutations}

% yshira20150421 I think the title of the section should change (because this is not "detail").

@@

This generative model is summarized in Fig. \ref{methodOverview}.
This model is essentially a ``mixed-membership model", also known as an ``admixture model" \cite{pmid10835412} or ``latent Dirichlet allocation" \cite{Blei:2003}. For example,
the membership proportions for each sample are analogous to admixture proportions in an admixture model; the mutation signatures are analogous to 
populations, and the mutation signature-specific parameters are analogous to population-specific allele frequencies.

The key parameters in this model
are the membership proportions for each sample, $\bm{q}_i$, and the mutation signature parameters, $\bm{F}_k$. 
We estimate these parameters by maximizing likelihood using an EM algorithm. 
A simulation study demonstrates that the estimation method can reproduce the mutation signature very accurately 
provided enough mutations and samples are available (see S1 Text). 
See Methods for more detailed models, parameter estimation, further discussion on relationships with mixed membership models, how to select $K$, etc.

\begin{figure}[h]
% \includegraphics[width=450pt]{methodOverview.pdf}
\caption{{\bf An overview of the generative model of somatic mutations proposed in this paper.} 
Suppose there are three types of mutation sources (mutation signatures) such as ultraviolet, tobacco smoking chemicals and transcription coupled repairs.
Each cancer genome has ratios showing which types of mutation sources are contributing to its mutations (membership parameters).
The generative model of the pattern of each mutation is:
first, one of the mutation signatures is chosen according to the membership parameter.
Second, each mutation feature such as substitution patterns and flanking bases is generated 
by the corresponding multinomial distributions for the selected mutation signature.
}
\label{methodOverview}
\end{figure}

@@

% Results and Discussion can be combined.
% \section*{Results}

@@

Here we compare our new ``independent model'' for 
mutation signatures, which assumes independence
among mutation features,
with the ``full model'', which corresponds to existing approaches. We compare mutation signatures obtained by the two approaches and investigate the robustness of each approach by down-sampling experiments.

@@

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and results in the next subsection (Figs. 3A and 3B).
In contrast, for the full model, this tendency is rather mild (Figs. 3C, 3D and S1).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

\begin{figure}[h]
\caption{{\bf The mutation signatures for the UCUT data, and the results of down-sampling experiments.}
3072 elements in the full model mutation signatures were shown divided by 6 substitution patterns and strand directions.
(A, B) APOBEC and AA signature for the independent model.
(C, D) APOBEC and AA signature for the full model.
(E, F) APOBEC and AA signature stability (the mean cosine similarity for each donw-sampling ratio).}
\label{UCUT}
\end{figure}

To investigate this further we performed down-sampling experiments.
Using the mutation signatures obtained using all 14717 substitutions as a gold standard, 
we assessed performance of the proposed method on down-sampled data consisting of $r\%$ of the original data, where  $r=(1\%, 2.5\%, 5\%, 10\%, 25\%, 50\%)$.
To measure robustness we used the cosine similarity on the 
full dimensional vector space, which allows comparison between the full model and the independent model.
We repeated each down-sampling experiment 100 times for each model.

@@

\begin{figure}[h]
\centering
% \includegraphics[width=16cm,height=17.5cm]{AlexandrovEtAl_mergedSignature.eps}
\caption{{\bf The summary of mutation signatures across 30 cancer types \cite{pmid23945592} obtained using the proposed method.}
Here, the substitution patterns and two 5' and 3' bases from the mutated sites are taken into account as mutation features.
First, mutation signatures were estimated separately in each cancer type, and then similar signatures were merged (see text).
}
\label{nature2013_sig_summary}
\end{figure}

@@

\begin{figure}[h]
% \subfigure[APOBEC signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=14.5cm,height=6cm]{APOBEC_two5prime.eps}
%    \label{APOBEC_two5prime}}
% \subfigure[POLE1 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE1_two5prime.eps}
%   \label{POLE1_two5prime}}
% \quad
% \subfigure[POLE2 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE2_two5prime.eps}
%   \label{POLE2_two5prime}}
% \subfigure[UV signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{UV_two5prime.eps}
%   \label{UV_two5prime}}
% \quad
% \subfigure[The signature 11 intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{LMST_two5prime.eps}
%   \label{LMST_two5prime}}
\caption{{\bf The estimated frequencies of bases at two 5' to the mutated site for each cancer type.}
The bar heights show the estimated frequency for bases A, C, G and T at two 5' to the mutated site.
The error bars show bootstrapped standard errors.
(A, B, C, D, E) The intensities of signature 13 (APOBEC signature), signature 1 (the first POL $\epsilon$ signature),
signature 8 (the second POL $\epsilon$ signature), signature 10 (ultraviolet signature) and signature 11, respectively, at two 5' to the mutated site.
intensities at two 5' to the mutated site.
}
\label{two5prime}
\end{figure}

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'', \cite{pmid23318258}). 
Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S3A and S3B), and signature 2 disappeared when we removed this sample (see Fig. S4).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

@@

We have detected a number of mutation signatures having transcription strand biases, 
which are naturally considered to be associated with transcription activities.
Therefore, to further understand the effect of transcription activities on mutational mechanisms, 
we can include gene expression or RNA polymerase II occupancies to mutation features, 
so that the relationships of strand biases and transcription activities will be clarified.

@@

We use the R package SQUAREM \cite{varadhan2008simple} to accelerate
convergence of this EM algorithm (SQUAREM implements a general approach to accelerate the convergence of any fixed-point iterative scheme such as an EM algorithm).
To address potential problems with convergence to local minima,
we apply the EM algorithm several times (10 times in this paper) using different initial points,
and use the estimate with the largest log-likelihood.
See S3 Text for the derivation of the above updating procedures.

@@

\subsection*{S1 Table}
\label{mfexample}
{\bf Example of representation for mutation patterns (substitution patterns and one 5' and 3' bases).}
In the independent representation, the elements of vector show substitution patterns, 5' adjacent bases and 3' adjacent bases, respectively.
For substitution pattens, 1 to 6 values are assigned to C$>$A, C$>$G, C$>$T, T$>$A, T$>$C and T$>$G in this order.
For 5' and 3' adjacent bases, 1 to 4 values are assigned to A, C, G and T.
Note that the original base is fixed to C or T to remove the redundancy of complement sequences.

\subsection*{S2 Table}
\label{nature2013_sigNum}
{\bf The number of mutation signatures selected for each cancer type for each cancer type in the Alexandrov et al. (2013) data.}

\subsection*{S1 Figure}
\label{UCUT_APOBEC_twoFivePrime}
{\bf The frequencies of bases at two 5' to the mutated site for the APOBEC mutation signatures obtained in UCUT data using the independent and full models.}

\subsection*{S2 Figure}
\label{nature2013_sig_list}
{\bf The list of several signatures extracted in each cancer type in the Alexandrov et al. (2013) data.}
(A) APOBEC signatures (signature 13) obtained in each cancer type.
(B) Smoking signature in each cancer type.
(C) The first Pol $\epsilon$ signature (signature 1)  in each cancer type.
(D) The second Pol $\epsilon$ signature (signature 8) in each cancer type.
(E) The ultraviolet signature i(signature 10) n each cancer type.
(F) Unknown signature (signature 11) obtained in lung small cell carcinomas and stomach cancers.

@@

\subsection*{S4 Figure}
\label{LGG_signatures}
{\bf The signatures obtained for the original data and for the data without the hyper-mutated case.}
(A) The result for the original data ($K = 3$). The first (from the left) signature seems to be one from deamination of 5-methyl-cytosine. 
The second signature is the low grade glioma signature.
(B, C, D, E) The result for the data without the hyper-mutated sample for $K = 2, 3, 4$ and $5$, respectively.
Although the signature related to deamination of 5-methyl-cytosine remained, low grade glioma specific signature could not be observed.

@@

We thank to Dr. Daichi Mochihashi for helpful discussion and comments on an earlier version of the method;
Dr. Hiromichi Suzuki for many helpful feedback on the R packages;
Dr. Ziyue Gao for reading the manuscript and providing many helpful comments.
